# NYC's Property Valuation and Assessment

## 9.1. Introduction

This project, devised from an interest in financial risk assessments, will explore New York City's Property Valuation and Assessment Data. This chapter displays the exploratory data analysis and visualization of the City's financial evaluations that determine tax revenues. Property assessments determine tax revenues, this identifying gaps between assessed property value and its market value is an important fiscal concern.

This project's goal is to encourage financial data transparency by generating accessible and easy to understand visualization that encourage public engagement with the city's finances and property valuation. A better New York is one with more informed stakeholders!

This chapter will engage with the following key questions:

1.  How do tax exemption programs influence property values?

2.  What is the relationship between tax classification and market value?

3.  How do tax exemption programs and property market value vary by borough?

4.  How does the valuation gap vary by tax classification and borough?

## 9.2. Data

```{r}
#| message: false
library(tidyverse)

PVA_data <- read_csv("https://data.cityofnewyork.us/resource/yjxr-fw8i.csv")

PVA_tidy <- PVA_data %>% 
  select(boro, taxclass, fullval, avtot, extot, exmptcl) %>% 
  mutate(across(c(boro, taxclass, exmptcl), as.factor))

str(PVA_tidy)
```

### 9.2.1 Description

The data used in this study is the NYC Property Valuation and Assessment Data, collected by the City government's Department of Finance (DoF), and sourced from [NYC Open Data](https://data.cityofnewyork.us/City-Government/Property-Valuation-and-Assessment-Data/yjxr-fw8i/about_data). It contains property valuation and assessment data for 9.85 million New York City properties, with a collection of 40 variables.

While the dataset is historic, updates occurred after its original publication. More importantly, due to the dataset's enormity, it is too large for spreadsheet programs like Excel and its downloading speed is slow. Thus, a `SODA2` API endpoint was used to export the data, in `CSV` format, ensuring the most up-to-date figures. The API limit for this data export was 1,000 rows.

The research conducted in this chapter is targeted, using a "tidied" version of the originally sourced dataset, containing the following variables:

+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| Variable Name  | Description                                                                                                                  | Data Type            |
+================+==============================================================================================================================+======================+
| **`BORO`**     | Classification of New York City's five boroughs.                                                                             | Factor (Nominal)     |
|                |                                                                                                                              |                      |
|                | -   **1** = Manhattan                                                                                                        |                      |
|                |                                                                                                                              |                      |
|                | -   **2** = Bronx                                                                                                            |                      |
|                |                                                                                                                              |                      |
|                | -   **3** = Brooklyn                                                                                                         |                      |
|                |                                                                                                                              |                      |
|                | -   **4** = Queens                                                                                                           |                      |
|                |                                                                                                                              |                      |
|                | -   **5** = Staten Island                                                                                                    |                      |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`TAXCLASS`** | Classification of property determining tax and assessment rates.                                                             | Factor (Ordinal)     |
|                |                                                                                                                              |                      |
|                | -   **1** = 1-3 Unit Residential                                                                                             |                      |
|                |                                                                                                                              |                      |
|                |     -   **1A** = One-Family Home                                                                                             |                      |
|                |                                                                                                                              |                      |
|                |     -   **1B** = Vacant Land (Residential)                                                                                   |                      |
|                |                                                                                                                              |                      |
|                |     -   **1C** = Condos (in 1-3 unit buildings)                                                                              |                      |
|                |                                                                                                                              |                      |
|                | -   **2** = Residential (4+ Units)                                                                                           |                      |
|                |                                                                                                                              |                      |
|                |     -   **2B** = 7-10 Unit Buildings                                                                                         |                      |
|                |                                                                                                                              |                      |
|                |     -   **2C** = 2-10 Unit Co-ops/Condos                                                                                     |                      |
|                |                                                                                                                              |                      |
|                | -   **3** = Utility Property                                                                                                 |                      |
|                |                                                                                                                              |                      |
|                | -   **4** = Commercial & All Other                                                                                           |                      |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`FULLVAL`**  | The property's market value.                                                                                                 | Numeric (Continuous) |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`AVTOT`**    | The property's assessed value, including the land any structures on it.                                                      | Numeric (Continuous) |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`EXTOT`**    | The total amount of the property's assessed value that is not taxable.                                                       | Numeric (Continuous) |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`EXMPTCL`**  | Classification for specific property tax exemption programs.                                                                 | Factor (Nominal)     |
|                |                                                                                                                              |                      |
|                | -   **X1**: NYC Government Owned (City agencies, parks, public schools).                                                     |                      |
|                | -   **X2**: NY State Government Owned (State offices, authorities).                                                          |                      |
|                | -   **X4**: Public Authorities (e.g., MTA, Port Authority, Housing Authority).                                               |                      |
|                | -   **X5**: Private Non-Profit / Public Benefit (Religious, charitable, or educational organizations that are fully exempt). |                      |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+

The data dictionary used to define the variables was incomplete, requiring further outside research on property and real estate terminology and classifications.

### 9.2.2 Missing value analysis

Describe any patterns you discover in missing values. If no values are missing, a graph should still be included showing that.

(suggested: 1-2 graphs plus commentary)

```{r}
library(redav)

# Missing values for each variable
colSums(is.na(PVA_tidy)) %>%
  sort(decreasing = TRUE)

# [GRAPH 1] Visualize Aggregated Missing Pattern Plot
plot_missing(PVA_tidy, percent = FALSE)

# Missingness by Borough and Property Type
missing_data <- PVA_tidy %>%
  group_by(boro, taxclass) %>%
  summarise(percent_na = mean(is.na(exmptcl)) * 100)

# [GRAPH 2] Visualize Missingness by Borough and Property Type
library(ggplot2)

ggplot(missing_data, 
       aes(x = percent_na, y = fct_reorder(taxclass, percent_na))) +
  geom_col() +
  facet_wrap(~boro)
```

The Aggregated Missing Pattern Plot indicates that the `exmptcl` variable is the source of missing values in my tidied dataset. 923 rows of this variable are missing. All the other variables are complete. The missing pattern in the `exmptcl` variable is likely systemic, occurring due to errors in data entry or cleaning.

The Missingness by Borough and Property Type Plot indicates that the missing data is high across all tax classes and borough, with percentages ranging from approximately 75% -100%. This is further proof of a systematic error, because the missingness is indiscriminate of borough or tax class.

## 9.3. Results

\[Provide a short summary of the most revealing findings of your analysis. Take extra care to clean up your graphs, ensuring that best practices for presentation are followed, as described in the audience ready style section below.

Your project should contain 5-10 graphs. Quality is rewarded over quantity. \]

### 9.3.1 Tax Exemption Programs and Property Values

### 9.3.2 Tax Classifications and Market Value

### 9.3.3 Market Value and Tax Exemption by Borough

### 9.3.4 Risk by Tax Classification and Borough

## 9.4. Conclusion

\[Discuss: main takeaways of your exploration, limitations, future directions, lessons learned.

(suggested length: one paragraph)\]
