# Food Security Data
Penny

## Introduction
Explain why you chose this topic, and the questions you are interested in studying. Provide context for readers who are not familiar with the topic.

(suggested length: one paragraph)

## Data
```{r}
library(tidyverse)
library(sf)
library(ggridges)
library(forcats)
library(redav)
library(tidycensus)
library(knitr)
library(scales)

food <- read.csv("data/Food/Food.csv")
Neighborhood <- read.csv("data/Food/2020_Neighborhood_Tabulation_Areas_(NTAs)_20260422.csv")

data <- Neighborhood |> 
  st_as_sf(wkt = "the_geom") |> 
  rename(geometry = the_geom) |> 
  left_join(food,
    by = c("NTA2020" = "Neighborhood.Tabulation.Area.NTA.")) |> 
  mutate(food_insecurity = parse_number(Food.Insecure.Percentage), 
         unemployment_rate = parse_number(Unemployment.Rate), 
         Year = as.factor(Year),
         Supply.Gap..lbs.. = parse_number(Supply.Gap..lbs..),
         unemployment_rate = case_when(
      Year %in% c("2023", "2024") ~ unemployment_rate * 100,
      Year == "2025" ~ floor(unemployment_rate))) |> 
  select(-c(`BoroCode`, `CountyFIPS`, `CDTA2020`, `NTAAbbrev`, `CDTAName`, 13, `Food.Insecure.Percentage`, `Unemployment.Rate`)) |>
  relocate(c(`Shape_Length`, `Shape_Area`,`geometry`), .after = last_col())

pl_2020 <- load_variables(2020, "pl", cache = T)

raw_race <- get_decennial(geography = "tract",
                           variables = c(`Total` = "P1_001N",
                                         `Hispanic or Latino` = "P2_002N",
                                         `Black-alone` = "P2_006N",
                                         `Asian-alone` = "P2_008N",
                                         `White-alone` = "P2_005N"),
                           state='NY',
                          county = c('New York', 'Kings', 'Bronx', 'Richmond',
                                     'Queens'),
                           geometry = T,
                           year = 2020,
                           output = "wide")

race <- raw_race %>% 
  mutate(`Hispanic or Latino (%)` = `Hispanic or Latino`/`Total`,
         `Black-alone (%)` = `Black-alone`/`Total`,
         `Asian-alone (%)` = `Asian-alone`/`Total`,
         `White-alone (%)` = `White-alone`/`Total`)

race <- raw_race %>% 
  mutate(`Hispanic or Latino (%)` = `Hispanic or Latino`/`Total`,
         `Hispanic or Latino (%)` = ifelse(is.nan(`Hispanic or Latino (%)`),
                                                NA, `Hispanic or Latino (%)`)) %>%
  mutate(`Black-alone (%)` = `Black-alone`/`Total`,
         `Black-alone (%)` = ifelse(is.nan(`Black-alone (%)`),
                                           NA, `Black-alone (%)`)) %>%
  mutate(`Asian-alone (%)` = `Asian-alone`/`Total`,
         `Asian-alone (%)` = ifelse(is.nan(`Asian-alone (%)`),
                                           NA, `Asian-alone (%)`)) %>% 
  mutate(`White-alone (%)` = `White-alone`/`Total`,
         `White-alone (%)` = ifelse(is.nan(`White-alone (%)`),
                                    NA, `White-alone (%)`))

data <- st_set_crs(data, 4326)

race_4326 <-  st_transform(race, 4326)

data_clean <- data |> select(NTAName, BoroName, geometry) |>
  distinct(NTAName, BoroName, .keep_all = TRUE)

race_nabes <- race_4326 |>
  st_join(data_clean, 
          left = TRUE, # left -> defines it as left_join -- meaning all census tract are kept
          join = st_intersects, # join -> defines the join definition as "if they intersect"
          largest = TRUE)

race_complete <- race_nabes |>
  st_drop_geometry()  |> select(-c(GEOID, NAME, BoroName)) |>
  group_by(NTAName) |> 
  summarise(Borough = first(BoroName),
            `Total` = sum(`Total`),
            `Total Hispanic or Latino` = sum(`Hispanic or Latino`),
            `Total Black-alone` = sum(`Black-alone`),
            `Total Asian-alone` = sum(`Asian-alone`),
            `Total White-alone` = sum(`White-alone`)) %>% 
  mutate(`Total Hispanic or Latino (%)` = percent(`Total Hispanic or Latino`/`Total`, accuracy = 1L),
         `Total Black-alone (%)` = percent(`Total Black-alone`/`Total`, accuracy = 1L),
         `Total Asian-alone (%)` = percent(`Total Asian-alone`/`Total`, accuracy = 1L),
         `Total White-alone (%)` = percent(`Total White-alone`/`Total`)) %>% 
  select(Borough, NTAName, Total, `Total Hispanic or Latino (%)`, `Total Black-alone (%)`,
         `Total Asian-alone (%)`, `Total White-alone (%)`)

data <- data |> left_join(race_complete, by = "NTAName")

head(data)
```
### Description
Describe how the data are collected and by whom. Describe the format of the data, the frequency of updates, dimensions, and any other relevant information. Note any issues / problems with the data, either known or that you discover.  Carefully document your sources with links to the precise data sources that you used. 

(suggested length: one paragraph)

### Missing value analysis
```{r}
data |> 
  rename (SuppGap = Supply.Gap..lbs.., VulnPopu = Vulnerable.Population.Score, WeigScor = Weighted.Score, FoodInsc = food_insecurity, EmplRate = unemployment_rate) |>
  plot_missing(num_char = 7, max_cols = 8, max_rows = 5)
data_missing <- data |> mutate(missing_value = is.na(Year))
data_missing |> ggplot() +
  geom_sf(aes(fill = missing_value)) +
  scale_fill_manual(
    values = c("FALSE" = "#B3B3B3","TRUE" = "#B49BE6")) + 
   geom_sf_text(data = data_missing |> filter(missing_value == TRUE), aes(label = NTAType), size = 1.5) + 
  labs(title = "Neighborhoods with Missing Values",
       x = "Longtitude", 
       y = "Latitude",
    fill = "Missing value",
    subtitle = "NTAType: Unlabeled = Residential; 5 = Rikers Island; 6 = Other Special Areas; 7 = Cemetery; 8 = Airport; 9 = Park")
```
Describe any patterns you discover in missing values. If no values are missing, a graph should still be included showing that.

(suggested: 1-2 graphs plus commentary)

## Results
### Question or topic 1
```{r}
data |> filter(Year %in% c("2023", "2024", "2025")) |> 
  ggplot() + geom_sf(aes(fill = food_insecurity)) +
  scale_fill_distiller(palette = "Greens", direction = 1,limits = c(0, 40), 
                       breaks = seq(0, 40, by = 10)) + 
  facet_wrap(~ Year, nrow = 1) +
  labs(title = "Food Insecurity Percentage by Neighborhood, 2023-2025", 
       x = "Longitude",
       y = "Latitude", 
       fill = "Food Insecurity Percentage") +
  theme(legend.position = "bottom")

data_diff <- data |> select(1:5, "Year", "food_insecurity") |> pivot_wider(names_from = Year, values_from = food_insecurity) |>
  mutate(`2023-2024` = `2024` - `2023`, `2024-2025` = `2025` - `2024`) |>
  pivot_longer(cols = c(`2023-2024`, `2024-2025`),
    names_to = "Year_diff",
    values_to = "food_diff")

data_diff |> ggplot() + geom_sf(aes(fill = `food_diff`)) +
  scale_fill_gradient2(low = "#2166AC", mid = "white",
                        high = "#B2182B", midpoint = 0, limits = c(-20, 20)) +
  facet_wrap(~ Year_diff, nrow = 1) +
  labs(title = "Change in Food Insecurity Percentage Points by Neighborhood", 
       x = "Longitude",
       y = "Latitude", 
       fill = "Change in % points")
```

### Question or topic 2
```{r}
p1 <- data |> filter (Year == "2025") |> group_by (BoroName) |> 
  ggplot(aes(x = food_insecurity, y = reorder(BoroName, food_insecurity, median))) + 
  geom_density_ridges(fill = "blue", alpha = 0.5, scale = 1) + 
  labs(x = "Food Insecurity Percentage", y = "Borough")

p2 <- data |> filter (Year == "2025") |> group_by(BoroName) |>
  ggplot(aes(x = food_insecurity, y = reorder(BoroName, food_insecurity, median))) +
  geom_boxplot() +
  xlim(c(0, 40)) + 
  labs(x = "Food Insecurity Percentage", y = "Borough")
p1 + p2 + plot_annotation(title = "Food Insecurity by Borough, 2025")

data |> ggplot(aes(x = Year, y = food_insecurity, group = NTAName)) +
  geom_line(alpha = 0.25) +
  geom_point(alpha = 0.25) +
  facet_wrap(~ BoroName) +
  labs(title = "Food Insecurity Trajectories by Borough", x = "Year", y = "Food Insecurity Percentage")

data |>  filter(Year == "2025") |> arrange(desc(food_insecurity)) |>
  mutate(NTA_Bor = paste0(NTAName, ",", BoroName)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = food_insecurity, y = reorder(NTAName, food_insecurity))) +
  geom_point(color = "blue") +
  labs(title = "Top 20 Neighborhoods by Food Insecurity, 2025",
    x = "Food Insecurity Percentage",
    y = "Neighborhood") 

```

### Question or topic 3
```{r}
data |> ggplot(aes(x = unemployment_rate, y = food_insecurity, color = Year)) +
  geom_jitter(width = 0.5, height = 0, alpha = 0.3) +
  geom_smooth(method = "lm",se = FALSE) + 
  labs(x = "Unemployment Rate",y = "Food Insecurity",color = "Year")

data <- data |> group_by(Year) |>
  mutate(scaled_vulpop = (Vulnerable.Population.Score-mean(Vulnerable.Population.Score))/sd(Vulnerable.Population.Score)) |>
  ungroup()

data |> ggplot(aes(x = scaled_vulpop, y = food_insecurity, color = Year)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Food Insecurity and Standardized Vulnerable Population Score",
    x = "Standardized Z Score for Vulnerable Population Score",
    y = "Food Insecurity",
    color = "Year")

data |> filter(!is.na(Year)) |> mutate(food_high = food_insecurity > median(food_insecurity, na.rm = TRUE),
    unemployment_high = unemployment_rate > median(unemployment_rate, na.rm = TRUE),
    category = case_when(
      food_high & unemployment_high ~ "High food insecurity, high unemployment", 
       food_high & !unemployment_high ~ "High food insecurity, low unemployment",
      !food_high & unemployment_high ~ "Low food insecurity, high unemployment",
      !food_high & !unemployment_high ~ "Low food insecurity, low unemployment")) |>
  ggplot() +
    facet_wrap(~ Year, nrow = 1)+
  geom_sf(aes(fill = category)) +
  labs(title = "Food Insecurity and Unemployment Classification, 2025",
       x = "Longtitude", 
       y = "Latitude",
    fill = "Category") + 
  theme(legend.position = "bottom")
```

## Conclusion
To be determined