# Bike New York City

Dany Rosete

## Introduction - CitiBike in NYC

## CitiBike Data 

### Description
The CitiBike dataset contains a detailed record of bike trips within NYC. Each row represents a completed ride. With this dataset, it is useful to analyze urban mobility as it includes geographical, time-based and user type data.

Here is the [link](https://citibikenyc.com/system-data) to the dataset. 

#### Structure of the Dataset 
This dataset is divided by Month directories that contain multiple `.csv` files. The complete dataset includes data from 2013 to March 2026 (as of May 3rd, 2026). 

Each row corresponds to a single bike trip that has the following fields: 

- Unique trip identifier (`ride_id`)

- Type of trip (`rideable_type`): electric vs classic bike

- Start time (`started_at`)

- End time (`ended_at`)

- Start station Name and identifier (`start_station_name` & `start_station_id`)

- End station name and identifier (`end_station_name` & `end_station_id`)

- Start Longitude & Latitude (`start_lng` & `start_lat`)

- End Longitude & Latitude (`end_lng` & `end_lat`)

- Type of Member (`member_casual`)

This dataset includes the following disclaimer:
"The data has been processed to remove (...) any trips that were below 60 seconds in length (potentially false starts or users trying to re-dock a bike to ensure it's secure)." 


#### Data Manipulation 

For this project, not all data will be used. Also, since .csv files are 100+ MB (exceeding the 100MB limit from Github), some files will be converted into Parquet format. 

To use Parquet files, `arrow` library is needed. An R script to convert the .csv files to parquet files is found in the /scripts folder. 

### Missing value analysis


```{r}
#| output: false
library(arrow)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(forcats)
library(redav)
```





### How Weather/Seasons affect use? 

```{r}
citibike_ds <- open_dataset("data/citibike_parquet")

seasonal <- citibike_ds |>
  mutate(
    hour = hour(started_at),
    wday = wday(started_at),
    month = month(started_at),
    date = as_date(started_at)) |> 
  count(month, wday, hour,date)|> 
  collect() |> 
  mutate(
    season = case_when(
      month %in% c(12,1,2) ~"Winter",
      month %in% c(3,4,5) ~ "Spring",
      month %in% c(6,7,8) ~ "Summer",
      month %in% c(9,10,11)~ "Fall"), 
    season = factor(season, levels = c("Winter", "Spring","Summer", "Fall")),
    wday = factor(wday, levels = 1:7,labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))) |>
  group_by(season, wday, hour) |>
  summarise(avg_rides = mean(n), .groups = "drop")

ggplot(seasonal, aes(x = hour, y = fct_rev(wday), fill = avg_rides)) +
  geom_tile(color = "white", linewidth = 0.1) +
  facet_wrap(~season) +
  scale_fill_viridis_c(option = "inferno") +
  scale_x_continuous(breaks= seq(0, 23, 6),
                     labels= paste0(seq(0, 23, 6),":00"),
                     expand = c(0, 0)) +
  labs(
    title = "Citi Bike rides by season (2025-2026)",
    x = "Hour of Day",
    y = NULL,
    fill = "Avgerage Rides") +
  theme_minimal(base_size = 12) +
  theme( legend.position  = "bottom", legend.key.width = unit(1.5, "cm"))
```


### Commuter or Leisure use? 





### Growth in outer boroughs ? ==NOT SURE==


## Conclusion 



